\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-01-31 analysis/paper/paper.tex
cbPaper draft for covars and interactions
draft paper for co-variates and interactions

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with a description
of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression.
\textcolor{red}{There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

@@

@@

@@

@@

\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}

@@

\begin{CodeOutput}
#> Call:
#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
#>     ScrArm, data = ERSPC)
#> 
#>   n= 159893, number of events= 540 
#> 
#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#>                       exp(coef) exp(-coef) lower .95 upper .95
#> ScrArmScreening group     0.801       1.25     0.674     0.952
#> 
#> Concordance= 0.519  (se = 0.011 )
#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
#> Wald test            = 6.37  on 1 df,   p=0.01
#> Score (logrank) test = 6.39  on 1 df,   p=0.01
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{../figures/erspc-cox-cif-1}

\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (B) Linear function of time which is equivalent to a Gompertz distribution. (C) The natural logarithm of time which is equivalent to a Weibull distribution. (D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{CodeChunk}

\hypertarget{equivalence-between-casebase-and-aft}{%
\subsection{Equivalence between casebase and
AFT}\label{equivalence-between-casebase-and-aft}}

@@

\begin{CodeOutput}
#>           (Intercept) ScrArmScreening group 
#>                 -7.77                 -0.22
\end{CodeOutput}

\begin{CodeOutput}
#>           (Intercept) ScrArmScreening group 
#>                 -7.76                 -0.23
\end{CodeOutput}

\begin{CodeOutput}
#>           (Intercept) ScrArmScreening group 
#>                -10.17                 -0.22
\end{CodeOutput}

\begin{CodeOutput}
#>           (Intercept)   log(Follow.Up.Time) ScrArmScreening group 
#>                 -9.44                  1.07                 -0.23
\end{CodeOutput}
\end{CodeChunk}

@@

@@

@@

@@

@@

@@

@@

\begin{CodeChunk}
\begin{table}[H]
\centering
\begin{tabular}{lrlr}
\toprule
Model & HR & CI & Width\\
\midrule
Cox & 0.80 & (0.67, 0.95) & 0.28\\
Exponential & 0.81 & (0.68, 0.96) & 0.28\\
Gompertz & 0.78 & (0.66, 0.93) & 0.27\\
Weibull & 0.80 & (0.67, 0.95) & 0.28\\
Splines & 0.81 & (0.68, 0.96) & 0.28\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
#> 1     54538       6053                                    
#> 2     54535       5786  3      267 <0.0000000000000002 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>    Exp.   Gomp.   Weib. Splines 
#>    6057    5816    6057    5796
\end{CodeOutput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
\centering
\begin{tabular}[t]{lll}
\toprule
Variable & Description & Statistical summary\\
\midrule
\code{Sex} & Sex & M=Male (100)\\
 &  & F=Female (77)\\
\code{D} & Disease & ALL (73)\\
 &  & AML (104)\\
\code{Phase} & Phase & CR1 (47)\\
\addlinespace
 &  & CR2 (45)\\
 &  & CR3 (12)\\
 &  & Relapse (73)\\
\code{Source} & Type of transplant & BM+PB (21)\\
 &  & PB (156)\\
\addlinespace
\code{Age} & Age of patient (years) & 4--62\\
 &  & 30.47 (13.04)\\
\code{ftime} & Failure time (months) & 0.13--131.77\\
 &  & 20.28 (30.78)\\
\code{Status} & Status indicator & 0=censored (46)\\
\addlinespace
 &  & 1=relapse (56)\\
 &  & 2=competing event (75)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{../figures/compPop-1}

\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study. A) The red points represent relapse events. B) The blue points represent the competing events.}\label{fig:compPop}
\end{figure}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.80 & (0.46, 1.39) & 0.68 & (0.39, 1.19)\\
Disease & 0.58 & (0.32, 1.03) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.16 & (0.46, 2.89) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.73 & (0.44, 6.72) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 3.93 & (1.82, 8.49) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.65 & (0.54, 5.03) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R+                           const(Phase) + const(Source) + const(Age), 
R+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{../figures/bmtcrr-risk-1}

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

@@

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within a hospital.
Imputation was conducted using predictive mean matching and using the
\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
there was a total of 1000 individuals and 35 variables. After
imputation, we have 474 individuals.
\textcolor{red}{How come we lose individuals during imputation?} Of
these, 67.5\% experienced the event of interest. A description of each
variable can be found in Table \ref{tab:support1} and a breakdown of
each categorical variable in Table \ref{tab:support2}. The original data
was obtained from the website of the Department of Biostatistics at
Vanderbilt University. The imputed dataset is available as part of the
\pkg{casebase} package.

@@

\begin{CodeInput}
R> support <- dplyr::select(support, -hospdead)
R> 
R> # Randomly select one observation
R> indiv <- sample(1:nrow(support), size = 1)
\end{CodeInput}
\end{CodeChunk}

Our main objective is to estimate the cumulative incidence of death for
a given covariate profile. Given the number of covariates available, we
will do this using regularized estimation. As \pkg{casebase} makes use
of the \pkg{glmnet} package, we will use the function
\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
Specifically, we input a matrix \code{y}, which contains the time and
event variables, and a matrix \code{x}, which contains all other
variables we would like to include in the model. For the SUPPORT
dataset, all factor variables were converted into dummy variables.

We are now ready to estimate the hazard function. We use lasso
penalization; this is achieved by setting \code{alpha} to 1. Also, note
that the argument \code{formula_time} can be used to specify the
function form for time; in our example, we model time linearly.

@@

\begin{CodeInput}
R> support_cb <- fitSmoothHazard.fit(support_train, y, 
R+                                   formula_time = ~ d.time, 
R+                                   family = "glmnet", 
R+                                   event = "death", time = "d.time", 
R+                                   alpha = 1, ratio = 100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model object and use the \code{absoluteRisk} function
to compute our estimate of the cumulative incidence function. Since
\code{glmnet} fitted the penalized logistic regression for a range of
penalty terms, we need to specify which one we want to use for our
prediction. We select \code{s = "lambda.1se"}, as recommended.

@@

\begin{CodeInput}
R> time_points <- seq(1, max(support$d.time), by = 0.5)
R> support_risk <- absoluteRisk(support_cb,
R+                              time = time_points,
R+                              newdata = support_test,
R+                              s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of Cox regression,
and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
can fit the Cox regression model using regularized estimation.

@@

\begin{CodeInput}
R> # Create survival object for glmnet
R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
R+                               event = y[, "death"])
R> 
R> support_cox <- glmnet::cv.glmnet(x = support_train, 
R+                                  y = surv_glmnet,
R+                                  family = "cox", alpha = 1)
\end{CodeInput}
\end{CodeChunk}

However, \pkg{glmnet} does not provide any functionality for estimating
the baseline hazard, and the functionalities in the \pkg{survival}
package do not work on the output from \code{cv.glmnet}. To circumvent
this issue and get an estimate of the baseline hazard, we followed the
following steps. First, we fitted a non-regularized Cox model using only
the variables selected by \pkg{glmnet}; the baseline hazard will be
estimated using this object. Then, we replaced the estimates of the
regression coefficients by the estimates from \pkg{glmnet}. Finally,
this modified \code{coxph} object is fed to the \code{survival::survfit}
function to obtain the estimated survival function.

@@

\begin{CodeInput}
R> km <- survival::Surv(time = support$d.time, 
R+                      event = support$death)
R> abKm <- survival::survfit(km ~ 1,
R+                           type = 'kaplan-meier')
\end{CodeInput}
\end{CodeChunk}

>>>>

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with some of the
functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation for ridge.
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
are the only packages to offer this functionality.
\textcolor{red}{I can dive deeper to determine if this is a fair statement}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets}, \pkg{cmprsk} and \pkg{survival} contain implementations for
competing risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression. Note that \pkg{CFC} has extra features
outside the scope of the metthods discussed in this paper.

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R>                        levels = c(0, 1), 
R>                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
R> plot(pt_object)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC, 
R>                                      event = "DeadOfPrCa", 
R>                                      exposure = "ScrArm")
R> plot(pt_object_strat)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R>                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R>                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R>        legend = c("Control group", "Screening group"), 
R>        col = c("red","blue"),
R>        lty = c(1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R>                                                   data = ERSPC, 
R>                                                   ratio = 100)
R> 
R> summary(casebase_exponential)
R> exp(coef(casebase_exponential)[2])
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R>                                  data = ERSPC, 
R>                                  ratio = 100)
R> 
R> summary(casebase_time)
R> exp(coef(casebase_time))
R> exp(confint(casebase_time))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R>                                     time = seq(0,15,0.1), 
R>                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> c("Exp." = AIC(casebase_exponential),
R>   "Weib." = AIC(casebase_time),
R>   "Splines" = AIC(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> 
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

@@

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
R>     data = bmtcrr, 
R>     ratio = 100, 
R>     time = "ftime")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R>                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R>                                      levels = levels(bmtcrr[,"Sex"])),
R>                       "D" = c("ALL", "AML"), # Both diseases
R>                       "Phase" = factor(c("Relapse", "Relapse"), 
R>                                        levels = levels(bmtcrr[,"Phase"])),
R>                       "Age" = c(35, 35),
R>                       "Source" = factor(c("PB", "PB"), 
R>                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R>                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R>                           const(Phase) + const(Source) + const(Age), 
R>                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within a hospital.
Imputation was conducted using the \pkg{mice} package in \proglang{R}
with default settings \citep{mice}. Before imputation, there were a
total of 1000 individuals and 35 variables. After imputation, we have
690 individuals. We lose individuals due to coliniarity. If a pair of
covariates are deemed colinear to a certain degree, they are excluded
from any imputations. Here, only edu income avtisst* race ph* glucose*
adlp* were imputed. Of these, 66.09\% experienced the event of interest.
A description of each variable can be found in Table \ref{tab:support1}
and a breakdown of each categorical variable in Table
\ref{tab:support2}. The original data was obtained from the Department
of Biostatistics at Vanderbilt University. The imputed dataset is
available as part of the \pkg{casebase} package.

@@

\begin{CodeInput}
R> xInteractions=x*y[,2]
R> colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
R> xFull=cbind(x,xInteractions)
R> cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
R+                                        formula_time = ~d.time,alpha=1,ratio=rat)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model and use the absoluteRisk function to integrate
and retrieve our absolute risk function.

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))
R> 
R> cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The cox regression absolute risk
curve with lasso penalization required two extra steps to retrieve.
Coxnet from the glmnet package has tools to fit a regularized cox model,
but requires that a survival object skeleton of the selected variables,
so that the survival package tools can be used to retrive the absolute
risk. We handle this by fitting a regularized cox model with the glmnet
package, and a cox model from the survival package with the selected
variables from the regularized fit. This second model serves as a
skeleton that permits the use of survival package functions. The
coefficients from the regularized model will replace the coeficients in
our skeleton. the resulting model will have the correct coefficients
when integrating, but will have an incorrect standard error. For the
purposes of this case study, we are only interested in the absolute risk
curve itself and not the standard error.

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> 
R> xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
R> colnames(xCoxI)[1]="(Intercept)"
R> 
R> #hazard for cox using glmnet
R> coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> #plot(coxNet)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
R> #The coefficients of this object will be replaced with the estimates from coxNet.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxNetFitI$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
R> #posCox=nonzero_covariate_cox$X1
R> #newDataCoxI=cleanCoxDataI[,posCox]
R> 
R> abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)
R> 
R> abCNI<-abcoxNetI
R> abCNI$surv=rowMeans(abCNI$surv)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
R> coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
R> abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
R> abCFI<-abcoxFitI
R> abCFI$surv<-rowMeans(abCFI$surv)
R> #cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
R>        #                                ratio=rat,data=fullDataCox)
R> #summary(cb.ModelNorm)#need to figure out how to deal with NAs
R> #casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
R> #cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
\end{CodeInput}
\end{CodeChunk}

We used the survival package to calculate the kaplan-meier absolute risk
function.

@@

\begin{CodeInput}
R> # A plot to compare all three Absolute risk (Commulative Incidence)
R> plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
R> lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
R> lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
R> legend("bottomright", 
R+        legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
R+        col = c("black","red","Blue","purple"),
R+        lty = c(1, 1, 1), 
R+        bg = "gray90")
\end{CodeInput}
\begin{figure}
